### 2023-12-21 sections/2-methods.tex
updated up to greedy sieve in methods

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number.
We require that the distance function be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
In addition to these constraints, if the distance function obeys the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
The choice of distance function varies by dataset.
For example, with neural-network embeddings, one could use Euclidean (L2) distance or cosine distance.
With biological or string data, Levenshtein or Hamming distances are more appropriate.
With time-series data, one could use dynamic time warping (DTW) distance.

Some of our algorithms for $\rho$- and $k$-nearest neighbors search rely on the \emph{local fractal dimension} at some point and length scale in the dataset.
We define local fractal dimension (LFD) as:

\begin{equation}
    \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) }
    \label{eq:methods:lfd_original}
\end{equation}

where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset.
To illustrate the difference, consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575.
Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws of physics that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional manifold within the 8,575-dimensional embedding space.
The \emph{local fractal dimension} is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
This notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space is known as the \emph{manifold hypothesis}~\cite{fefferman2016testing}.

% Najib: Add our own plots of local fractal dimension instead of referencing figures from the CHESS paper. Then, rewrite this paragraph.
Figures 2 and 3 in~\cite{ishaq2019clustered} illustrate the low fractal dimensions of two datasets used with $\rho$-NN search in CHESS. 
We observe that for both APOGEE (Figure 2 in~\cite{ishaq2019clustered}) and GreenGenes (Figure 3 in~\cite{ishaq2019clustered}), other than the most extreme 10\% of clusters, virtually all clusters have a local fractal dimension of less than 2. 
This suggests that APOGEE and GreenGenes are good datasets for use with entropy-scaling search algorithms like CHESS and CAKES.

@@

Najib We define a \emph{cluster} as a set of points with a \emph{center} and a \emph{radius}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance of any point in the cluster to the \emph{center}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
We estimate local fractal dimension at the scale of the cluster radius and half the cluster radius. 
This allows us to rewrite Equation \ref{eq:methods:lfd_original} as simply: 
\begin{equation} 
\log_2 \frac{|B_X(q, r)}{|B_X(q, r/2)|}
\end{equation}

% Najib: Rewrite this paragraph. Old paper defined it as ... but we define it for the hierarchical clustering scheme as ... DONE. (let me know if this okay)

In~\cite{yu2015entropy}, metric entropy for a given cluster radius $r_c$ was defined as the number of clusters of radius $r_c$ needed to cover all data. In this paper, we define the metric entropy $\mathcal{N}_{\hat{r_c}}(X)$ of a data set $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree.

@@

We start by building a divisive hierarchical clustering of the dataset using CLAM.
The procedure is similar to what was outlined in CHESS, but with the following improvements:
better selection of poles for partitioning and depth-first dataset reordering (see Section ~\ref{subsubsec:methods:clustering:dataset-depth-first-reordering}).

CAKES assumes the manifold hypothesis. 
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$. 
While we sometimes use Euclidean notions, such as voids and volumes, to talk about geometric and topological properties of clusters and of the manifold, CLAM does not rely on such notions; 
they serve merely as a convenient, intuitive way to talk about the underlying mathematics.

@@

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in the sample.
Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
i.e., the point which minimizes the sum of distances to all other points in the sample.
This geometric median is $C$'s \emph{center}.
The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $\ell$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}.
The point $r$ which is furthest from $\ell$ is designated the \emph{right pole}.
We then partition the cluster into a \emph{left child} and \emph{right child}, where the left child contains all points in the cluster which are closer to $\ell$ than to $r$, and the right child contains all points in the cluster which are closer to $r$ than to $\ell$.
Without loss of generality, we assign to the left child those points which are equidistant from $\ell$ and $r$.
Starting from a root-cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion is met.
This process is described in Algorithm \ref{alg:methods:clustering:partition}.

\begin{algorithm} % enter the algorithm environment
\caption{Partition(\emph{C})} % give the algorithm a caption
\label{alg:methods:clustering:partition} % and a label for \ref{} commands later in the document
\begin{algorithmic} % enter the algorithmic environment
    \STATE $m \Leftarrow \lfloor \sqrt{|C|} \rfloor$
    \STATE $seeds \Leftarrow m$ random points from $C.points$
    \STATE $c \Leftarrow$ geometric median of $seeds$
    \STATE $l \Leftarrow \argmax d(c,x) \ \forall \ x \in C.points$
    \STATE $r \Leftarrow \argmax d(l,x) \ \forall \ x \in C.points$
    \STATE $left \Leftarrow \{x | x \in C.points \land d(l,x) \le d(r,x)\}$
    \STATE $right \Leftarrow \{x | x \in C.points \land d(r,x) < d(l,x)\}$
    \IF{$|left| > 1$}
        \STATE Partition($left$)
    \ENDIF
    \IF{$|right| > 1$}
        \STATE Partition($right$)
    \ENDIF
\end{algorithmic}
\end{algorithm}

@@

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.
In CHESS, each cluster stored a list of indices into the dataset, which we used during search to retrieve a cluster's points. 
Though this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high;
since each cluster stores indices for each of its points and, for a dataset with cardinality $n$, we store a total of $n$ indices at each depth in the tree.
Assuming a balanced tree and thus $\mathcal{O}\log(n)$ depth, this approach had $\mathcal{O}(n\log(n))$ memory cost.

One may improve this memory cost to $\mathcal{O}(n)$ by only storing these indices at the leaf cluster level.
This approach, however, introduces an $\mathcal{O}(n\log(n))$ time cost whenever we need to find the indices for 
a non-leaf cluster, since it requires a traversal of the subtree rooted at that cluster.

In this work, we introduce a new approach, wherein after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store the cluster's cardinality and an \emph{offset} to access the cluster's points from the dataset.
The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.
With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers significant improvements over the approach used in CHESS.

@@

The asymptotic complexity of Partition is the same as described in~\cite{ishaq2019clustered}
By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of partitioning and $\mathcal{O}(n \log n)$ cost of building the tree.
This is a significant improvement over exact partitioning and tree-building, which have $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ costs respectively.

>>>>

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number.
We require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$) and have an identity (i.e. $f(x, y) = 0 \iff x = y$). 
In addition to these constraints, if the distance function obeys the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
The choice of distance function varies by dataset.
For example, with neural-network embeddings, one could use Euclidean (L2) distance or Cosine distance.
With biological or string data, Levenshtein or Hamming distances are more appropriate.
With time-series data, one could use Dynamic Time Warping (DTW).

Some of our search algorithms rely on the \emph{local fractal dimension} (LFD) at some point and length scale in the dataset.
We define LFD as:

\begin{equation}
    \frac{\text{log}(\frac{|B_X(q, r_1)|}{|B_X(q, r_2)|})}{\text{log}(\frac{r_1}{r_2}) }
    \label{eq:methods:lfd-original}
\end{equation}

where $B_X(q, r)$ is the set of points contained in a ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
We stress that this concept differs from the \emph{embedding dimension} of a dataset.
To illustrate the difference, consider SDSS's APOGEE dataset, wherein each datum is a nonnegative real-valued vector of length 8,575.
Hence, the \emph{embedding dimension} of this dataset is 8,575. 
However, due to physical constraints (namely, the laws of physics that govern stellar fusion and spectral emission lines), the data are constrained to a lower-dimensional manifold within the 8,575-dimensional embedding space.
The LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
This is the \emph{manifold hypothesis}~\cite{fefferman2016testing}, i.e. high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space.

@@

We define a \emph{cluster} as a set of points with a \emph{center} and a \emph{radius}.
The \emph{center} is the geometric median of the points (or a smaller sample of the points) in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance from the \emph{center} to any point in the \emph{cluster}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
We estimate LFD at the scale of the cluster radius and half that radius. 
This allows us to rewrite Equation~\ref{eq:methods:lfd-original} as:

\begin{equation} 
    \log_2 \left( \frac{|B_X(q, r)}{|B_X(q, r/2)|} \right)
    \label{eq:methods:lfd-simplified}
\end{equation}

In~\cite{yu2015entropy}, metric entropy for a given cluster radius $r_c$ was defined as the number of clusters of a uniform radius $r_c$ needed to cover all data.
In this paper, we define the metric entropy $\mathcal{N}_{\hat{r_c}}(X)$ of a data set $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree.

@@

We start by performing a divisive hierarchical clustering on the dataset using CLAM.
The procedure is similar to what was outlined in CHESS, but with the following improvements:
better selection of poles for partitioning (see Algorithm~\ref{alg:methods:clustering:partition}) and depth-first reordering of the dataset (see Section~\ref{subsubsec:methods:clustering:dataset-depth-first-reordering}).

CAKES assumes the manifold hypothesis. 
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$. 
While we sometimes use Euclidean notions, such as voids and volumes, to talk about the geometric and topological properties of clusters and of the manifold, CLAM does not rely on such notions; 
they serve only as a convenient, intuitive way to talk about the underlying mathematics.

@@

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in this sample.
Using these distances, we compute the \emph{geometric median} of this sample; 
i.e., the point which minimizes the sum of distances to all other points in the sample.
This geometric median is the \emph{center} of $C$.

The \emph{radius} of $C$ is the maximum distance from the \emph{center} to any other point in $C$.

The point which is responsible for that radius (i.e., the furthest point from the \emph{center}) is designated the \emph{left pole}.
The point which is furthest from \emph{left pole} is designated the \emph{right pole}.

We then partition the cluster into a \emph{left child} and a \emph{right child}, where the left child contains all points in the cluster which are closer to the \emph{left pole} than to the \emph{right pole}, and the right child contains all points in the cluster which are closer to the \emph{right pole} than to the \emph{left pole}.
Without loss of generality, we assign to the left child those points which are equidistant from the two poles.

Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria.
This process is described in Algorithm \ref{alg:methods:clustering:partition}.

During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-simplified}.

\begin{algorithm} % enter the algorithm environment
    \caption{Partition(\emph{C})} % give the algorithm a caption
    \label{alg:methods:clustering:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $C$, a cluster
        \STATE $seeds \Leftarrow \lfloor \sqrt{|C|} \rfloor$ random points from $C$
        \STATE $c \Leftarrow$ geometric median of $seeds$
        \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$
        \IF{$|L| > 1$}
            \STATE Partition($L$)
        \ENDIF
        \IF{$|R| > 1$}
            \STATE Partition($R$)
        \ENDIF
    \end{algorithmic}
\end{algorithm}

@@

CAKES also improves upon CHESS by reordering the dataset in depth-first order of traversal of the tree.

In CHESS, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Though this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree.
Assuming a balanced tree, and thus $\mathcal{O}\log(n)$ depth, this approach had a memory overhead of $\mathcal{O}(n\log(n))$.

One may improve this memory overhead to $\mathcal{O}(n)$ by only storing indices at the leaf clusters.
This approach, however, introduces a time cost of $\mathcal{O}(n\log(n))$ whenever we need to find the indices for a non-leaf cluster (because it requires a traversal of the subtree rooted at that cluster).

In this work, we introduce a new approach wherein, after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \emph{cardinality} and an \emph{offset} to access the its points from the dataset.

The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset as that of its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.

With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers significant improvements over the approach used in CHESS.

@@

The asymptotic complexity of Partition is the same as described in~\cite{ishaq2019clustered}.
By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of partitioning and $\mathcal{O}(n \log n)$ cost of building the tree.
This is a significant improvement over exact partitioning and tree-building, which cost $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ respectively.

\subsection{The Search Problem}
\label{subsec:methods:the-search-problem}

Given Algorithm~\ref{alg:methods:clustering:partition}, we can now pose the $k$-NN and $\rho$-NN search problems.

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find the set $S_q$ such that $|S_q| = k$ and $\forall x \in \textbf{X} \setminus S_q$, $\ f(x, q) > \max\{f(y, q): y \in S_q \}$;
i.e., for a given $k$, find the $k$ closest points to $q$ in $\textbf{X}$.
We also have the $\rho$-nearest neighbors search problem, which aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$.

\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta^{+} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
    \item $\delta^{-} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
\end{itemize}

% Najib: For this figure, I want the red, green and blue lines/braces to be thinner. Perhaps 1/2 to 2/3 as thick. Also, I changed the notation for the deltas because it makes things easier in the Sieve and Greedy Sieve algorithms' sections.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.png}
    \caption{$\delta$, $\delta^{+}$, and $\delta^{-}$ for a cluster $C$ and a query $q$.}
    \label{fig:methods:deltas}
\end{figure}

We define a \emph{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains duplicates of the same point (i.e., has cardinality greater than 1 but contains only one \emph{unique} point).
A singleton clearly has zero radius, and so $\delta = \delta^{-} = \delta^{+}$.
Hence, we overload the above notation to also refer to the distance from a query to an individual point.

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which might contain points in the query ball.

% Najib: Visually, ``m'' is not centered on the line segment. It's a little bit to the right. Moving ``l'' to the right is probably the easiest fix.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-3.pdf}
    \caption{
        $m$ denotes the midpoint of $\overline{l r}$. $q'$ is the projection of $q$ onto $\overline{l r}$.
        $d$ is the distance from $q'$ to $m$.
        If $d \leq \rho$, both child clusters could contain points in the query ball.
        Otherwise, only the right child cluster could contain such points. 
    }
    \label{fig:methods:overlapping-children-3}
\end{figure}

To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children-3}.
Here, we overload the notation for $\overline{x y}$ to refer to the line segment joining points $x$ and $y$ as well as the length of that line segment.

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles (see Section~\ref{subsubsec:methods:cluster-partitioning}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$ (i.e., the query is closer to $r$ than to $l$).

Now let $q'$ be the projection of $q$ to $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$ and $d$ be the distance from $q'$ to $m$.
Since we assign a point in the parent cluster to the left child if it is closer to the left pole, if $d > \rho$ then the left child cannot contain points inside the query ball.
In this case, we proceed only with the right child.
If, however, $d \leq \rho$ then both child clusters could contain points in the query ball, and so we proceed with both of them.

To check whether $d \leq \rho$, we note that

\begin{equation}
    d = \overline{m q' \vphantom{l}} = \overline{m r \vphantom{l}} - \overline{q' r \vphantom{l}} = \frac{\overline{l r}}{2} - \overline{q' r \vphantom{l}}.
    \label{eq:methods:d-expression}
\end{equation}

Let $\theta$ denote $\angle l r q$, as shown in Figure~\ref{fig:methods:overlapping-children-3}.
By the Law of Cosines on $\triangle l r q$, we have that

\begin{equation}
    \text{cos}(\theta) = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r} \cdot \overline{q r \vphantom{l}}}.
    \label{eq:methods:cosine-law}
\end{equation}

Since $\triangle r q q'$ is a right triangle, we also have that

\begin{equation}
    \text{cos}(\theta) = \tfrac{\overline{q' r \vphantom{l}}}{\overline{q r \vphantom{l}}}.
    \label{eq:methods:cosine-law-right-triangle}
\end{equation}

Combining Equations~\ref{eq:methods:cosine-law} and~\ref{eq:methods:cosine-law-right-triangle}, we have that

\begin{equation}
    \overline{q' r \vphantom{l}} = \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}}.
    \label{eq:methods:q-prime-r}
\end{equation}

Substituting Equation~\ref{eq:methods:q-prime-r} into Equation~\ref{eq:methods:d-expression}, we have that

\begin{equation*}
    d = \tfrac{\overline{l r}}{2} - \tfrac{\overline{q r \vphantom{l}}^2 + \ \overline{l r}^2 - \ \overline{q l}^2}{2 \cdot \overline{l r}} = \tfrac{\overline{q l}^2 - \overline{q r \vphantom{l}}^2}{2 \cdot \overline{l r}}.
\end{equation*}

Thus,

\begin{equation}
    d \leq \rho \iff (\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho.
    \label{eq:methods:check-d}
\end{equation}

Note, in particular, that Equation~\ref{eq:methods:check-d} only requires distances between real points, and so it can be used with any distance function, even when $q'$ and $m$ are not real points or cannot be imputed from the data.

To perform $\rho$-nearest neighbors search, we first perform a coarse tree-search, as outlined in Algorithm~\ref{alg:methods:rnn:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained search, as outlined in Algorithm~\ref{alg:methods:rnn:leaf-search}, to find all points in which are within a distance $\rho$ of the query.

\begin{algorithm} 
    \caption{tree-search($C$, $q$, $\rho$)} 
    \label{alg:methods:rnn:tree-search} 
    \begin{algorithmic}
        \REQUIRE $C$, a cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \STATE $c$ $\Leftarrow$ \emph{center} of $C$
        \STATE $r$ $\Leftarrow$ \emph{radius} of $C$

        \IF{$C$ is a leaf \textbf{or} $f(q, c) + r \leq \rho$}
            \STATE \textbf{return} $\{C\}$
        \ELSIF{Inequality~\ref{eq:methods:check-d} is true}
            \STATE $L$ $\Leftarrow$ \emph{left child} of $C$
            \STATE $R$ $\Leftarrow$ \emph{right child} of $C$
            \STATE \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
        \ELSIF{$f(q, l) < f(q, r)$}
            \STATE \textbf{return} tree-search($L, q, \rho$)
        \ELSE
            \STATE \textbf{return} tree-search($R, q, \rho$)
        \ENDIF
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{leaf-search($S$, $q$, $\rho$)} 
    \label{alg:methods:rnn:leaf-search} 
    \begin{algorithmic}
        \REQUIRE $S$, a set of clusters
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \STATE $H \Leftarrow \emptyset$, a set of hits

        \FOR{$C \in S$}
            \STATE $c$ $\Leftarrow$ \emph{center} of $C$
            \STATE $r$ $\Leftarrow$ \emph{radius} of $C$
            \IF{$f(q, c) + r \leq \rho$}
                \STATE $H$ $\Leftarrow$ $H \cup \{C\}$
            \ELSE
                \FOR{$p \in C$}
                    \IF{$f(p, q) \leq \rho$}
                        \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                    \ENDIF
                \ENDFOR
            \ENDIF
        \ENDFOR

        \STATE \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)} 
    \label{alg:methods:rnn:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \STATE $S$ $\Leftarrow$ tree-search($root$, $q$, $\rho$)
        \STATE $H$ $\Leftarrow$ leaf-search($S$, $q$, $\rho$)
        \STATE \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{\log_2 \mathcal{N}_{\hat{r}}(X)}_{\textrm{metric entropy}} +
        \overbrace{ \left| B_X(q, \rho) \right|}^{\textrm{output size}}
        \underbrace{ \left( 1 + \frac{2 \cdot \hat{r}}{ \rho} \right) ^ d}_{\textrm{scaling factor}}
    \Bigg)
    \label{eq:methods:rnn:complexity}
\end{gather}

where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query in a region of radius $r$.
From Algorithm~\ref{alg:methods:rnn:rnn-search}, we have that $H = B_X(q, \rho)$, and so $\rho$-nearest neighbors search performance scales linearly with the size of the output set and exponentially with the local fractal dimension around the query.
While one might worry that the exponential scaling factor would dominate, the manifold hypothesis suggests that the local fractal dimension around the query is typically very low, and so this algorithm actually scales sub-linearly with the cardinality of the dataset.

In this section, we present three novel algorithms for exact $k$-nearest neighbors search:
Repeated $\rho$-NN, Sieve, and Greedy Sieve.

In these algorithms, we use $H$, for \emph{hits}, to refer to the data structure which stores the closest points to the query found so far and $Q$ to refer to the data structure which stores the clusters and points which are still in contention for being one of the $k$ nearest neighbors.

\subsubsection{Repeated \texorpdfstring{$\rho$}{p}-NN}
\label{subsubsec:methods:knn-search:repeated-rnn}

In this algorithm, we perform $\rho$-nearest-neighbors search starting with a small search radius, and repeatedly increasing the radius until $|H| \geq k$.

Let the search radius, $r$, be equal to the radius of the root cluster divided by the cardinality of the dataset.
We perform tree-search with radius $r$.
If no clusters are returned, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.
Let $S$ be the set of clusters returned by the first tree-search which returns at least one cluster.

Now, so long as $\sum_{C \in S} |C| < k$, we continue to perform tree-search, but instead of doubling $r$ on each iteration, we multiply it by a factor determined by the LFD in the vicinity of the query ball. 
In particular, we increase the radius by a factor of

\begin{equation}
    \min \left(2, \left( {\frac{k}{\sum_{C \in S} |C|}} \right)^{\mu^{-1}} \right)
    \label{eq:repeated-rnn-factor}
\end{equation}

where $\mu$ is the harmonic mean of the LFD of the clusters in $S$.
We use the harmonic mean to ensure that $\mu$ is not dominated by outlier clusters with very high LFD. 
Intuitively, the factor by which we increase the radius should also be \emph{inversely} related to the number of points found so far. 
Additionally, when the LFD at the radius scale from the previous iteration is low, this suggests that the data are relatively concentrated in that region.
In such a region, a small increase in the radius would likely encounter unoccupied space, so a larger radial increase is needed.
Thus, the factor of radius increase should be \emph{inversely} related to the LFD.

Once $\sum_{C \in S} |C| >= k$, we are guaranteed to have found at least $k$ neighbors, and so we can stop increasing the radius.
We perform $\rho$-nn-search with radius $r$ and return the $k$ nearest neighbors.

\begin{algorithm} % enter the algorithm environment
    \caption{repeated-$\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:knn:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find

        \STATE $r \Leftarrow radius$ of the $root$ cluster
        \STATE $r \Leftarrow$ $\frac{r}{|root|}$
        \STATE $S \Leftarrow \emptyset$, a set of clusters

        \WHILE {$S = \emptyset$}
            \STATE $S \Leftarrow$ tree-search($root$, $q$, $r$)
            \STATE $r \Leftarrow 2 \cdot r$
        \ENDWHILE

        \ALOOP{}
            \IF{$\sum_{C \in S} |C| >= k$}
                \STATE \textbf{break}
            \ENDIF
            \STATE $\mu \Leftarrow \frac{|S|}{\sum_{C \in S} \frac{1}{LFD(C)}}$
            \STATE $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in S} |C|}} \right)^{\mu^{-1}} \right)$
            \STATE $S \Leftarrow$ tree-search($root$, $q$, $r$)
        \ENDLOOP

        \STATE $H \Leftarrow$ sort($\rho$-NN-search($root$, $q$, $r$))
        \STATE \textbf{return} $H[.. k]$
    \end{algorithmic}
\end{algorithm}

\subsubsection{Complexity of Repeated \texorpdfstring{$\rho$}{p}-NN}
\label{subsubsec:methods:repeated-rnn-complexity}

The complexity bounds for Repeated $\rho$-NN rely on the assumption that the query point is sampled from the same distribution as the rest of the data, i.e. it arises from the same generative process as the rest of the data.
Given the uses of $k$-NN search in practice, this assumption is reasonable.
From this assumption, we can infer that the LFD at the query point does not differ significantly from the (harmonic) mean LFD of clusters near the query.

We find it useful to adopt the terminology used in \cite{ishaq2019clustered} and \cite{yu2015entropy}, and address \emph{tree-search} and \emph{leaf-search} separately.
Tree-search refers to the process of identifying clusters which have overlap with the query ball, i.e., which might contain one of the $k$ nearest neighbors. 
Leaf-search refers to the process of identifying the $k$ nearest neighbors among the points in those clusters identified by tree-search.

In ~\cite{ishaq2019clustered}, we showed that the complexity of tree-search is $\mathcal{O}(\log\mathcal{N}_{\hat{r}}(X))$, where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset $X$ at a radius $\hat{r}$.
To adjust this bound for Repeated $\rho$-NN, we must estimate the number of iterations of tree-search needed to find at least $k$ neighbors.

Based on the assumption that the LFD at the query point is not significantly different from that of nearby clusters, the factor from Equation~\ref{eq:repeated-rnn-factor} suggests that in the expected case, we need only two iterations of tree-search to find $k$ neighbors:
one iteration to find at least one neighbor, and the one more to find enough remaining neighbors.
Since this is a constant factor, it does not affect the asymptotic complexity of Repeated $\rho$-NN.

To determine the asymptotic complexity of leaf-search, we must estimate $\sum_{C \in S} |C|$, the number of points we examine.
Let $\rho_k$ be the distance from the query to the $k$th nearest neighbor.
Then, we see that $S$ is expected to be the set of clusters which overlap with a ball of radius $\rho_k$ around the query.
We can estimate this region as a ball of radius $\rho_k + \hat{r}$, where $\hat{r}$ is the mean radius of clusters in $S$.

\begin{equation*}
    \sum_{C \in S} |C| \leq \gamma  \left| B(q, \rho_k) \right| \left(1+ \frac{2\hat{r}}{\rho_k} \right)^d
\end{equation*}

where $\gamma$ is a constant. 
By definition of $\rho_k$, we have that $|B(q, \rho_k)| = k$.
Thus,

\begin{equation}
    \sum_{C \in S} |C| \leq \gamma k \left( 1+ \frac{2\hat{r}}{\rho_k} \right)^d
    \label{eq:methods:repeated-rnn-leaf-search}
\end{equation}

It remains to provide an estimate for $\rho_k$. 
For this, we once again rely on the assumption that the query is from the same distribution as the rest of the data, and thus the LFD at the query point is not significantly different than that of nearby clusters.

We let $\hat{d}$ be the mean LFD of the clusters in $S$.
While ordinarily we compute LDS by comparing cardinalities of two balls with two different radii centered at \emph{the same} point, in order to estimate $\rho_k$, we instead compare the cardinality of a ball \emph{around the query} of radius $\rho_k$ to the mean cardinality, $\hat{C}$, of clusters in $S$ at a radius equal to the mean of their radii, $\hat{r}$.

We justify this approach by noting that, since the query is from the same distribution as the rest of the data, we could replace the query with the center of one of the nearby clusters without significantly changing the LFD at the query point.

By Equation~\ref{eq:repeated-rnn-factor}, we then have that

\begin{equation*}
    \hat{d} = \frac{\log{}\frac{\hat{|C|}}{k}}{\log{}\frac{\hat{r}}{\rho_k}}
\end{equation*}

Since $\hat{d}$, $\hat{C}$, $\hat{r}$, and $k$ are all known values, we can solve Equation~\ref{eq:methods:repeated-rnn-leaf-search} for $\rho_k$ to get

\begin{equation*}
    \rho_k = \hat{r} \left( \frac{k}{\hat{C}} \right)^{\hat{d}^{-1}}
\end{equation*}

Simplifying the term for leaf search in Equation~\ref{eq:methods:rnn:complexity}, we get

\begin{equation*}
    k \left( 1+ \frac{2\hat{r}}{\rho_k} \left( \frac{k}{\hat{C}} \right) ^ {\hat{d}^{-1}} \right)^d
\end{equation*}

Since we are performing an asymptotic analysis, we can ignore the additive factor of $1$.
By combining the bounds for tree-search and leaf-search, we see that Repeated $\rho$-NN has an asymptotic complexity of
